Technique for \earlywarning\ filtering for compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-29 introduction.tex
Section 1.3: this is still not in strict chronological order, and I think that readers can infer that it is in loosely chronological order, so we no longer explicitly state that it is
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1193 ccdd05d5-56c1-461a-bf98-85f6876523b4
Tuned latency budget in section 1.3
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1194 ccdd05d5-56c1-461a-bf98-85f6876523b4

There were a number of sources of latency associated with \CBC{}
analysis in S6/VSR3 \citep{HugheyGWPAW2011}, listed here in chronological order.
%
\begin{itemize}
%
\item {\it Data aggregation and broadcasting ($\sim$$0.25$\,s):}
The \LIGO{} data acquisition system records science
data in 1/16\,s blocks \citep{Bork2001}. Data must also be aggregated from all
of the \GW\ observatories using global computer networks capable of high
bandwidth but perhaps only modest latency.  This could introduce an additional
latency of $\sim$100\,ms.  These latencies are much smaller than the
other sources of latency below and would require significant reengineering and
capital investments to overcome.
%
\item {\it Data calibration, quality control, and distribution
($\sim$$2$\,min):} There is technical latency in calibrating the detector data,
assessing its quality, and distributing it to a central location for analysis.
None of this latency is fundamental and improvements in other areas could make
this the dominating source of latency, so work should be done to improve this.
%
\item {\it Trigger production ($2$--$5$\,min):} This was an impressive latency
and was certainly sub-dominant in the S6/VSR3 analysis.  However, even if no
other sources of latency exist, triggering latency is too long to catch prompt
or extended
emission. It is also more challenging for searches with advanced detectors
because the templates needed to filter the data are about ten times longer.  We
will focus on reducing the latency of trigger production in this work.
%
\item {\it Event gathering, downselection, and pointing calculations
($2$--$3$\,min):} S6/VSR3 saw the introduction of low-latency astronomical
alerts, which required gathering event parameters and sky localization from the
various online analyses, downselecting the events, and calculating pointings.
If other sources of latency improve, this could dominate, so work should be
done to improve this.
%
\item {\it Human vetting of the trigger quality ($10$--$20$\,min):} In
commissioning the new alert system in S6/VSR3, humans manually examined each
candidate for obvious flaws before publishing an alert. This was by far the
largest latency during S6/VSR3 and hopefully confidence will grow to the
point where human operators are satisfied with checks after alerts have been
transmitted. We give it no further consideration here.
%
\end{itemize}

>>>>

There were a number of sources of latency associated with \CBC{}
analysis in S6/VSR3 \citep{HugheyGWPAW2011}, listed here.
%
\begin{itemize}
%
\item {\it Data acquisition and aggregation ($\gtrsim$100~ms):}
The \LIGO\ data acquisition system collects data from detector subsystems 16
times a second~\citep{Bork2001}. Data are also copied from all of the \GW\
observatories to the analysis clusters over the Internet, which is capable of
high bandwidth but perhaps only modest latency.  Together, these introduce a
latency of $\gtrsim 100$~ms.  These technical sources of latency could be reduced
with significant engineering and capital investments, but they are minor compared
to any of the other sources of latency.
%
\item {\it Data conditioning
($\sim$1~min):} Science data must be calibrated using the detector's frequency
response to gravitational radiation.  Currently, data are calibrated in blocks of
16~s.  Within $\sim$1~min, data quality is assessed in order to create veto flags.
These are both technical sources of latency that might be addressed with improved
calibration and data quality software for advanced detectors.
%
\item {\it Trigger generation (2--5~min):} Low-latency data analysis pipelines
deployed in S6/VSR3 achieved an impressive latency of minutes.  However, second
to the human vetting process, this dominated the latency of the entire \EM\
followup process.  Even if no other sources of latency existed, this trigger
generation latency is too long to catch prompt or even extended emission.
Low-latency trigger generation will become more challenging with advanced detectors
because inspiral signals will stay in band up to ten times longer.  In this work,
we will focus on reducing this source of latency.
%
\item {\it Alert generation
(2--3~min):} S6/VSR3 saw the introduction of low-latency astronomical
alerts, which required gathering event parameters and sky localization from the
various online analyses, downselecting the events, and calculating pointings.
If other sources of latency improve, the technical latency associated with this
infrastructure could dominate, so work should be done to improve it.
%
\item {\it Human validation (10--20~min):} Because the new alert system was
commissioned during S6/VSR3, humans examined each candidate for obvious flaws
before publishing an alert. This was by far the largest latency during S6/VSR3.
Hopefully, confidence in the system will grow to the point where no human intervention
is necessary before alerts are sent, so we give it no further consideration here.
%
\end{itemize}
